---
title: 'Homework: Numerical linear algebra'
output:
  html_notebook:
    code_folding: none
    highlight: textmate
---

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: {
    Macros: {
      bA: "{\\boldsymbol{A}}",
      bx: "{\\boldsymbol{x}}",
      bb: "{\\boldsymbol{b}}"
    }
  }
});
</script>


```{r setup, include=FALSE}
source(file.path("..", "R", "util.R"))

required_packages <- c("microbenchmark")
install_and_load_packages(required_packages)
```


# Exercise 1: Comparing different numerical linear algebra algorithms for solving linear systems

In this exercise, we consider the problem of solving a linear system $\bA \bx = \bb$ for $\bx$.
We compare the three methods we learned in the class: LU, Cholesky, and QR decompositions.
(Of course, LU applies to more general systems and QR can be used to solve least squares, but here we focus on positive definite systems.)

## Part A: Racing for solution &mdash; speed comparison 

We first compare their computational speed. 
Fill in the code below and `bench::mark()` the three algorithms.

**Questions:**
What are relative speeds among the algorithms?
Do relative speeds agree with what you expect from the complexity analyses?
If not (quite), why might that be?

**Note:**
I misspoke about the R's built-in `chol()` function during the lecture:
when applied to a positive-definite `A`, the function actually returns an _upper-triangular_ matrix `R` such that `t(R) %*% R == A`.

**Answer:**
The speed in the solve/Cholesky is about twice of QR, which is quite different from the complexity analysis, as QR is $4/3p^3$ when n=p (in our case) while Cholesky is $p^3$.

```{r, eval=FALSE}
# Import the `rand_positive_def_matrix` function
source(file.path("R", "random_matrix.R"))

mat_size <- 1024L
A <- rand_positive_def_matrix(mat_size, cond_num = 1e3)
b <- rnorm(mat_size)

solve_via_cholesky <- function(A, b) {
  ## solve "Ax = b" question via cholesky decomp
  L <- chol(A)
  X <- backsolve(t(L), forwardsolve(L, t(A) %*% b))
  return(c(X))
}

bench::mark(
  solve(A, b)
)
bench::mark(
  solve_via_cholesky(A, b)
)
bench::mark(
  qr.solve(A, b)
)
```

## Part B: Competition in terms of numerical accuracy/stability

We now compare the three methods in terms of numerical accuracy/stability.
To this end, we set up the following simulation study. 
We first generate a "ground truth" solution vector $\bx_0$.
We then compute an "numerical-error-free" $\bb = \bA \bx_0$ by carrying out the matrix-vector multiplication using the `long double` type, which (on most hardware and compilers) provides [additional 12 bits of precision](https://en.wikipedia.org/wiki/Extended_precision#x86_extended_precision_format).
Of course, the vector $\bb$ computed as such still suffers from numerical errors, but the idea is that the numerical errors from this high-precision matrix-vector multiplication is much smaller than the errors caused by numerically solving $\bA \bx = \bb$ for $\bx$.
We can thus assess the accuracy of the three solvers by comparing the numerically computed $\bx$ to the ground truth $\bx_0$.

### Task &#x2F00;

First compare the outputs of matrix-vector multiplication $\bx \to \bA \bx$ using `double` and `long double` using the provided Rcpp functions.

**Questions:**
What is the relative difference in $\ell^2$-norm? 
How about the coordinate-wise relative differences?
Are the observed magnitudes of the differences what you'd expect?

**Answer:**
Both the $\ell^2$-norm and the coordinate-wise relative differences are in the scale of e-16, which is as expected as the long double provides more accuracy when rounding the numbers.

```{r, eval=FALSE}
Rcpp::sourceCpp(file.path("src", "matvec_double.cpp"))
Rcpp::sourceCpp(file.path("src", "matvec_ldouble.cpp"))

set.seed(1918)

mat_size <- 1024L
A <- rand_positive_def_matrix(mat_size, cond_num = 1e3)
x <- rnorm(mat_size)

# Fill in
double <- matvec_double(A, x)
ldouble <- matvec_ldouble(A, x)

l2 <- sqrt(sum((double - ldouble)^2)) / sqrt(sum(double^2))
coordinate_wise_diff <- abs(double - ldouble) / abs(double)
```

### Task &#x2F06;

Now randomly generate $\bA$ so that its condition number is $10^6$.
Then solve a positive-definite system $\bA \bx = \bb$ for $\bx$ using the three algorithms and compare their outputs to the ground truth $\bx_0$.

**Questions:**
Which algorithm appears to be more accurate than the others? 
Visually demonstrate your answer.

**Answer**With only visual check, the QR and the solve() is more accurate than cholesky.

```{r, eval=FALSE}
set.seed(1918)
cond_num <- 1e6

# Larger matrices could incur substantial computational time under base R BLAS
mat_size <- 1024L 

A <- rand_positive_def_matrix(mat_size, cond_num)
x <- rnorm(mat_size)

# Fill in
b <- matvec_ldouble(A, x)

x_solve <- solve(A, b)
x_chol <- solve_via_cholesky(A, b)
x_qr <- qr.solve(A, b)

df <- data.frame(x = x, x_solve, x_chol, x_qr)
df_noChol <- data.frame(x = x, x_solve, x_qr)

# Convert data to long format
df_long <- tidyr::gather(df, key = "group", value = "y", -x)
df_long_noChol <- tidyr::gather(df_noChol, key = "group", value = "y", -x)
library(ggplot2)
# Create plot with jittered scatter plot
ggplot(df_long, aes(x = x, y = y, color = group)) +
  geom_jitter(width = 0.2, height = 0.2) +
  theme_classic() +
  labs(x = "x",
       y = "y",
       color = "Group")

ggplot(df_long_noChol, aes(x = x, y = y, color = group)) +
  geom_jitter(width = 0.2, height = 0.2) +
  theme_classic() +
  labs(x = "x",
       y = "y",
       color = "Group")
```

### Task &#x4E09;

In Task &#x2F06;, we compared the three algorithms in one randomly generated example.
Now we consider a more systematic (though hardly comprehensive) comparison via repeated simulations.
We also vary the condition number of $\bA$ and assess whether the results would hold across varying degrees of ill-conditioning.

**Questions/To-do's:**

* Using the starter code provided, calculate various summary measures of the numerical errors.
* Integrate into the provided code one another (or more, if you like) meaningful metric(s) of your choice to summarize the numerical error.
* Visually explore how the three algorithms compare with each other in their accuracy. See if different error metrics tell different stories; they might or might not.
* Vary the condition number in the range $10^6 \sim 10^{12}$, e.g. by trying $10^6$, $10^9$, and $10^{12}$.
* Do you see any patterns in the numerical errors across the three algorithms, metrics, and/or condition numbers? Show some plots to support your conclusion.

**Answers**
When comparing three algorithms in all the metrics fixing the condition number, the method of `lu` and `qr` outperforms `cholesky` in all metrics describing accuracies. For the `cholesky`, while the normalized and 5th/95th errors vary a lot among different rounds of simulations, the medians appears to be much stable, and the errors within each rounds are low.

Changing the condition number, the errors in cholesky becomes even bigger.

**Note:** 
The QR solver will throw an error when the system is ill-condition enough that the numerical solution might not be very accurate. 
To force it to return the solution in any case, set `tol = .Machine$double.eps`.

```{r, eval=FALSE}
# Utility functions for bookkeeping simulation results.
source(file.path("R", "num_linalg_sim_study_helper.R"))

n_sim <- 32L
mat_size <- 512L
cond_num <- 1e6
metrics <- c("norm", "median", "five_percentile", "ninety_five_percentile",
             "variance")
  # TODO: add another metric and modify the helper script accordingly

rel_error_list <- lapply(
  c("lu", "chol", "qr"), 
  function(method) pre_allocate_error_list(n_sim, metrics)
)

for (sim_index in 1:n_sim) {
  A <- rand_positive_def_matrix(mat_size, cond_num)
  x <- rnorm(mat_size) 
  b <- matvec_ldouble(A, x)
  x_approx <- list( 
    lu = solve(A, b),
    chol = solve_via_cholesky(A, b),
    qr = qr.solve(A, b, tol = .Machine$double.eps)
  )
  for (method in c("lu", "chol", "qr")) {
    rel_error <- lapply(
      metrics, 
      function (metric) calc_rel_error(x, x_approx[[method]], metric)
    )
    names(rel_error) <- metrics
    for (metric in names(rel_error)) {
      rel_error_list[[method]][[metric]][sim_index] <- rel_error[[metric]]
    }
  }
}


## comparing norm
library(tidyverse)
data.frame(lu = rel_error_list[["lu"]][["norm"]],
           chol = rel_error_list[["chol"]][["norm"]],
           qr = rel_error_list[["qr"]][["norm"]],
           rounds = 1:32) %>%
  pivot_longer(cols = c("lu", "chol", "qr"),
               values_to = "error_norm",
               names_to = "method") %>%
  ggplot(aes(x=rounds, y=error_norm, color=method)) +
    geom_point()
## comparing median
data.frame(lu = rel_error_list[["lu"]][["median"]],
           chol = rel_error_list[["chol"]][["median"]],
           qr = rel_error_list[["qr"]][["median"]],
           rounds = 1:32) %>%
  pivot_longer(cols = c("lu", "chol", "qr"),
               values_to = "error_median",
               names_to = "method") %>%
  ggplot(aes(x=rounds, y=error_median, color=method)) +
    geom_point()
## comparing ninety_five_percentile
data.frame(lu = rel_error_list[["lu"]][["ninety_five_percentile"]],
           chol = rel_error_list[["chol"]][["ninety_five_percentile"]],
           qr = rel_error_list[["qr"]][["ninety_five_percentile"]],
           rounds = 1:32) %>%
  pivot_longer(cols = c("lu", "chol", "qr"),
               values_to = "error_ninety_five_percentile",
               names_to = "method") %>%
  ggplot(aes(x=rounds, y=error_ninety_five_percentile, color=method)) +
    geom_point()


## comparing five_percentile

data.frame(lu = rel_error_list[["lu"]][["five_percentile"]],
           chol = rel_error_list[["chol"]][["five_percentile"]],
           qr = rel_error_list[["qr"]][["five_percentile"]],
           rounds = 1:32) %>%
  pivot_longer(cols = c("lu", "chol", "qr"),
               values_to = "error_five_percentile",
               names_to = "method") %>%
  ggplot(aes(x=rounds, y=error_five_percentile, color=method)) +
    geom_point()

## comparing variance
data.frame(lu = rel_error_list[["lu"]][["variance"]],
           chol = rel_error_list[["chol"]][["variance"]],
           qr = rel_error_list[["qr"]][["variance"]],
           rounds = 1:32) %>%
  pivot_longer(cols = c("lu", "chol", "qr"),
               values_to = "error_variance",
               names_to = "method") %>%
  ggplot(aes(x=rounds, y=error_variance, color=method)) +
    geom_point()

```
